---
title: "GEO880: Exercises Week 2"
author: "Antonia Hehli"
date: "2023-04-28"
output: html_document
---

# Task 1 - Import your data

```{r import libraries}
library(dplyr)
library(ggplot2)
library(readr) 
library(sf)
library(tmap)
library(zoo)
```


```{r import the data}
# read the data
wildschwein_BE <- read_delim("wildschwein_BE_2056.csv", ",")

# convert it to a spatial object
wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), 
                           crs = 2056, 
                           remove = FALSE)
```

# Task 2 - Getting an overview

Calculate the time difference between subsequent rows as described in the demo. You can calculate the time difference using the function difftime() in combination with lead().

- The function difftime() has an option units. Set this to secs to get the time difference in seconds
use as.integer() to turn the output returned by difftime() into an integer. Store the output in a new column (e.g. timelag)

```{r calculate time difference}
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"))

```

Now inspect your data in more detail. Try to answer the following questions:

- How many individuals were tracked?

```{r individuals}
wildschwein_BE_count <- wildschwein_BE %>%
  group_by(TierID) %>%
  count()

print(unique(wildschwein_BE$TierName))
```

There are three individuals that were tracked. They have the IDs 002A, 0016A, and 018A and are called Sabi, Rosa, and Ruth.

- For how long were the individual tracked? Are there gaps? Were all individuals tracked concurrently or sequentially?

```{r plot observation time}
ggplot(wildschwein_BE, aes(DatetimeUTC, TierName)) + 
  geom_point()
```

Sabi was tracked from about September 2014 to August 2015; Ruth from about November/December 2014 until August 2015. And Rosa was tracked from also about November/December 2014 to July 2015. The individual were mostly tracked concurrently.

- What is the temporal sampling interval between the locations?

```{r plot histogram}
# timelag/60 = minutes
ggplot(wildschwein_BE, aes(timelag/60)) + 
  geom_histogram(binwidth = 0.5) + 
  lims(x = c(0, 5000/60))+
  labs(x = "Time lag [min]",
       y = "Count")
```

```{r plot the line of the temporal sampling}
wildschwein_BE %>%
  filter(DatetimeUTC < "2014-08-24") %>%
  ggplot(aes(DatetimeUTC, timelag, color = TierName)) + 
  geom_point() + 
  geom_line() +
  labs(x = "Date and Time [UTC]",
       y = "Time lag")
```

# Task 3 - Deriving movement parameters I: Speed

```{r calculation of step length with test values}
# save the first two sets of coordinates to find a good way to calculate the step length [m]
N1 <- 1204752
N2 <- 1204863
E1 <- 2570409
E2 <- 2570402

# calculate the speed
sqrt((E1-E2)^2 + (N1-N2)^2)
```

```{r calculate step length}
# use mutate() and lead() to calculate the step length [m]
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2))
```

```{r calculate speed}
wildschwein_BE <- wildschwein_BE %>% 
  mutate(speed_ms = steplength_m/timelag)
```

```{r speed histogram}
# plot the speed histogram using ggplot
ggplot(wildschwein_BE, aes(speed_ms)) + 
  geom_histogram(binwidth = 0.005)
```

# Task 4 - Cross-scale movement analysis

```{r read caro data}
# read the caro data
caro <- read_delim("caro60.csv", ",")
```

```{r prepare and slice the caro data}
# compute the total number of observations and save this value as a variable
length_caro <- nrow(caro)

# create the different sequences for the caro data
seq_caro3 <- seq(from = 1, to = length_caro, by = 3)
seq_caro6 <- seq(from = 1, to = length_caro, by = 6)
seq_caro9 <- seq(from = 1, to = length_caro, by = 9)

# create the caro subsamples using the created sequences
caro_3 <- dplyr::slice(caro, seq_caro3)
caro_6 <- dplyr::slice(caro, seq_caro6)
caro_9 <- dplyr::slice(caro, seq_caro9)

```

```{r calculate steplength and speed for each caro data set}
# for the initial data set: calculate steplength and speed using the formulas from task 3
caro <- caro %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)

# repeat that three times with the three different sequences
caro_3 <- caro_3 %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)

caro_6 <- caro_6 %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)

caro_9 <- caro_9 %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)
```

```{r plot speeds as line}
ggplot() + 
  geom_line(data = caro, aes(DatetimeUTC, speed_ms, color = "#00BFC4")) +
  geom_line(data = caro_3, aes(DatetimeUTC, speed_ms, color = "#7CAE00")) +
  geom_line(data = caro_6, aes(DatetimeUTC, speed_ms, color = "#C77CFF")) + 
  geom_line(data = caro_9, aes(DatetimeUTC, speed_ms, color = "#F8766D"))+
  labs(x = "Time",
       y = "Speed [m/s]",
       title = "Comparing derived speed at different sampling intervals")+
  scale_color_manual(name = "Sampling interval",labels = c("1 Minute", "3 Minutes",
                                               "6 Minutes", "9 Minutes"),
                     values=c("#C77CFF", "#7CAE00", "#00BFC4", "#F8766D"))+
  theme_minimal()
```

```{r plot the different trajectories of the caro data}
# combine the 1 and 3 minute trajectory into one data set
caro_1_3 <- rbind(caro, caro_3)

caro_1_3 %>% 
  group_by(as.factor(timelag_s)) %>% 
  ggplot(aes(x = E, y = N, color = as.factor(timelag_s)))+
  geom_point()+
  geom_path()+
  scale_color_discrete(na.translate = FALSE, name = "Trajectory [min]", labels = c("1 Minute", "3 Minutes"))+
  labs(title = "Comparing original- with 3 minutes-resampled data")+
  theme_minimal()

# repeat for one and six minutes
caro_1_6 <- rbind(caro, caro_6)

caro_1_6 %>% 
  group_by(as.factor(timelag_s)) %>% 
  ggplot(aes(x = E, y = N, color = as.factor(timelag_s)))+
  geom_point()+
  geom_path()+
  scale_color_discrete(na.translate = FALSE, name = "Trajectory [min]", labels = c("1 Minute", "6 Minutes"))+
  labs(title = "Comparing original- with 6 minutes-resampled data")+
  theme_minimal()

# repeat for one and nine minutes
caro_1_9 <- rbind(caro, caro_9)

caro_1_9 %>% 
  group_by(as.factor(timelag_s)) %>% 
  ggplot(aes(x = E, y = N, color = as.factor(timelag_s)))+
  geom_point()+
  geom_path()+
  scale_color_discrete(na.translate = FALSE, name = "Trajectory [min]", labels = c("1 Minute", "9 Minutes"))+
  labs(title = "Comparing original- with 9 minutes-resampled data")+
  theme_minimal()
```

# Task 5 - Deriving movement parameters II: Rolling window function

For this task, I needed to install a new packages called "zoo". After installing, I added it to the libraries at the beginning of this document.

```{r familiarize with the package}
# create a random example with 10 values
example <- rnorm(10)

# example with k = 3 as width of the rolling window
rollmean(example, k = 3, fill = NA, align = "left")

# example with k = 4 as width of the rolling window
rollmean(example, k = 4, fill = NA, align = "left")

```

```{r rollmean on caro}
# calculate the rolling mean
caro <- caro %>%
  mutate(speed_rm3 = rollmean(speed_ms, k = 3, fill = NA, align = "left"),
         speed_rm5 = rollmean(speed_ms, k = 5, fill = NA, align = "left"),
         speed_rm7 = rollmean(speed_ms, k = 7, fill = NA, align = "left"))

# plot the speeds with different rolling means
ggplot(data = caro) + 
  geom_line(aes(DatetimeUTC, speed_ms, color = "#00BFC4")) +
  geom_line(aes(DatetimeUTC, speed_rm3, color = "#7CAE00")) +
  geom_line(aes(DatetimeUTC, speed_rm5, color = "#C77CFF")) + 
  geom_line(aes(DatetimeUTC, speed_rm7, color = "#F8766D"))+
  labs(x = "Time",
       y = "Speed [m/s]")+
  scale_color_manual(name = "Speed",labels = c("Regular speed", "Rolling window (k = 3)",
                                               "Rolling window (k = 5)", "Rolling window (k = 7)"),
                     values=c("#C77CFF", "#7CAE00", "#00BFC4", "#F8766D")) +
  theme_minimal()
```

# Task 7 - Explore your movement data

## Import the data and convert it to a sf object

```{r load the movement data}
# read the data
movement <- read_delim("private_files/posmo_tracks.csv", ",")

# convert the movement data to a spatial object, the CRS is WGS84
movement <- st_as_sf(movement, coords = c("lon_x", "lat_y"), 
                           crs = 4326, 
                           remove = FALSE)

```

## Convert the CRS to 1903+ LV95

```{r transform CRS}
movement <- movement %>% 
  st_transform(crs = 2056)
```

## Create a map

```{r plot all the dots colored by transportation mode}
tmap_mode("view")
tm_shape(movement, projection = 2056)+
  tm_dots(col = "transport_mode")
```

## Speed calculation

```{r}
# get the LV95 coordinates in two separate coordinates
movement <- cbind(movement, st_coordinates(movement))

# calculate the speed
movement <- movement %>%
  mutate(timelag_s = as.integer(difftime(lead(datetime), datetime),
                           units = "secs"),
         steplength_m = sqrt((X-lead(X))^2 + (Y-lead(Y))^2),
         speed_ms = steplength_m/timelag_s)

# plot a histogram of the speeds by transportation modes
ggplot(movement, aes((speed_ms)*3.6, fill = transport_mode))+ # multiply m/s * 3.6 to get km/h
  geom_histogram(binwidth = 1)+
  lims(x = c(0, 300))+
  labs(x = "Speed [km/h]",
       y = "Count",
       title = "Speed histogram by transportation mode")+
  theme_minimal()+
  scale_fill_discrete(name = "Transport mode", 
                      na.translate = FALSE) # do not show NA values in histogram

```






