---
title: "GEO880: Exercises Week 2"
author: "Antonia Hehli"
date: "2023-04-28"
output: html_document
---

# Task 1 - Import your data

```{r import libraries}
library(dplyr)
library(ggplot2)
library(readr) 
library(sf)
library(zoo)
```


```{r import the data}
# read the data
wildschwein_BE <- read_delim("wildschwein_BE_2056.csv", ",")

# convert it to a spatial object
wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), 
                           crs = 2056, 
                           remove = FALSE)
```

# Task 2 - Getting an overview

Calculate the time difference between subsequent rows as described in the demo. You can calculate the time difference using the function difftime() in combination with lead().

- the function difftime() has an option units. Set this to secs to get the time difference in seconds
use as.integer() to turn the output returned by difftime() into an integer.

- store the output in a new column (e.g. timelag)

```{r calculate time difference}
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"))

```

Now inspect your data in more detail. Try to answer the following questions:

- How many individuals were tracked?

```{r individuals}
wildschwein_BE_count <- wildschwein_BE %>%
  group_by(TierID) %>%
  count()

print(unique(wildschwein_BE$TierName))
```

There are three individuals that were tracked. They have the IDs 002A, 0016A, and 018A and are called Sabi, Rosa, and Ruth.

- For how long were the individual tracked? Are there gaps? Were all individuals tracked concurrently or sequentially?

```{r plot observation time}
ggplot(wildschwein_BE, aes(DatetimeUTC, TierName)) + 
  geom_point()
```

Sabi was tracked from about September 2014 to August 2015; Ruth from about November/December 2014 until August 2015. And Rosa was tracked from also about November/December 2014 to July 2015. The individual were mostly tracked concurrently.

- What is the temporal sampling interval between the locations?

```{r plot histogram}
# timelag/60 = minutes
ggplot(wildschwein_BE, aes(timelag/60)) + 
  geom_histogram(binwidth = 0.5) + 
  lims(x = c(0, 5000/60))+
  labs(x = "Time lag [min]",
       y = "Count")
```

```{r plot the line of the temporal sampling}
wildschwein_BE %>%
  filter(DatetimeUTC < "2014-08-24") %>%
  ggplot(aes(DatetimeUTC, timelag, color = TierName)) + 
  geom_point() + 
  geom_line() +
  labs(x = "Date and Time [UTC]",
       y = "Time lag")
```

# Task 3 - Deriving movement parameters I: Speed

```{r calculation of step length with test values}
# save the first two sets of coordinates to find a good way to calculate the step length [m]
N1 <- 1204752
N2 <- 1204863
E1 <- 2570409
E2 <- 2570402

# calculate the speed
sqrt((E1-E2)^2 + (N1-N2)^2)
```

```{r calculate step length}
# use mutate() and lead() to calculate the step length [m]
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2))
```

```{r calculate speed}
wildschwein_BE <- wildschwein_BE %>% 
  mutate(speed_ms = steplength_m/timelag)
```

```{r speed histogram}
# plot the speed histogram using ggplot
ggplot(wildschwein_BE, aes(speed_ms)) + 
  geom_histogram(binwidth = 0.005)
```

# Task 4 - Cross-scale movement analysis

```{r read caro data}
# read the caro data
caro <- read_delim("caro60.csv", ",")
```

```{r prepare and slice the caro data}
# compute the total number of observations and save this value as a variable
length_caro <- nrow(caro)

# create the different sequences for the caro data
seq_caro3 <- seq(from = 1, to = length_caro, by = 3)
seq_caro6 <- seq(from = 1, to = length_caro, by = 6)
seq_caro9 <- seq(from = 1, to = length_caro, by = 9)

# create the caro subsamples using the created sequences
caro_3 <- dplyr::slice(caro, seq_caro3)
caro_6 <- dplyr::slice(caro, seq_caro6)
caro_9 <- dplyr::slice(caro, seq_caro9)

```

```{r calculate steplength and speed for each caro data set}
# for the initial data set: calculate steplength and speed using the formulas from task 3
caro <- caro %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)

# repeat that three times with the three different sequences
caro_3 <- caro_3 %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)

caro_6 <- caro_6 %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)

caro_9 <- caro_9 %>%
  mutate(timelag_s = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"),
         steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2),
         speed_ms = steplength_m/timelag_s)
```

```{r plot speeds as line}
# still try to add a legend
ggplot() + 
  geom_line(data = caro, aes(DatetimeUTC, speed_ms), color = "#F8766D") +
  geom_line(data = caro_3, aes(DatetimeUTC, speed_ms), color = "#7CAE00") +
  geom_line(data = caro_6, aes(DatetimeUTC, speed_ms), color = "#00BFC4") + 
  geom_line(data = caro_9, aes(DatetimeUTC, speed_ms), color = "#C77CFF")+
  labs(x = "Time",
       y = "Speed [m/s]")+
  theme_minimal()
```

```{r plot the different trajectories of the caro data}
ggplot()+ 
  geom_point(data = caro, aes(E, N), color = "#F8766D", alpha = 0.5)+
  geom_path(data = caro, aes(E, N),color = "#F8766D", alpha = 0.5) + 
  geom_point(data = caro_3, aes(E, N), color = "#00BFC4") + 
  geom_path(data = caro_3, aes(E, N), color = "#00BFC4") + 
  theme_minimal()

ggplot()+ 
  geom_point(data = caro, aes(E, N), color = "#F8766D", alpha = 0.5)+
  geom_path(data = caro, aes(E, N),color = "#F8766D", alpha = 0.5) + 
  geom_point(data = caro_6, aes(E, N), color = "#00BFC4") + 
  geom_path(data = caro_6, aes(E, N), color = "#00BFC4")+
  theme_minimal()

ggplot()+ 
  geom_point(data = caro, aes(E, N), color = "#F8766D", alpha = 0.5)+
  geom_path(data = caro, aes(E, N),color = "#F8766D", alpha = 0.5) + 
  geom_point(data = caro_9, aes(E, N), color = "#00BFC4") + 
  geom_path(data = caro_9, aes(E, N), color = "#00BFC4")+
  theme_minimal()
```

# Task 5 - Deriving movement parameters II: Rolling window function

For this task, I needed to install a new packages called "zoo". After installing, I added it to the libraries at the beginning of this document.

```{r familiarize with the package}
# create a random example with 10 values
example <- rnorm(10)

# example with k = 3 as width of the rolling window
rollmean(example, k = 3, fill = NA, align = "left")

# example with k = 4 as width of the rolling window
rollmean(example, k = 4, fill = NA, align = "left")

```

```{r rollmean on caro}
caro <- caro %>%
  mutate(speed_rm3 = rollmean(speed_ms, k = 3, fill = NA, align = "left"),
         speed_rm5 = rollmean(speed_ms, k = 5, fill = NA, align = "left"),
         speed_rm7 = rollmean(speed_ms, k = 7, fill = NA, align = "left"))

ggplot(data = caro) + 
  geom_line(aes(DatetimeUTC, speed_ms, color = "#00BFC4")) +
  geom_line(aes(DatetimeUTC, speed_rm3, color = "#7CAE00")) +
  geom_line(aes(DatetimeUTC, speed_rm5, color = "#C77CFF")) + 
  geom_line(aes(DatetimeUTC, speed_rm7, color = "#F8766D"))+
  labs(x = "Time",
       y = "Speed [m/s]")+
  scale_color_manual(name = "Speed",labels = c("Regular speed", "Rolling window (k = 3)",
                                               "Rolling window (k = 5)", "Rolling window (k = 7)"),
                     values=c("#C77CFF", "#7CAE00", "#00BFC4", "#F8766D"))
```

# Task 6 - Add your movement data to your repository



