---
title: "GEO880: Exercises Week 2"
author: "Antonia Hehli"
date: "2023-04-28"
output: html_document
---

# Task 1 - Import your data

```{r import the data}
library(ggplot2)
library("readr") 
library("sf") 

# read the data
wildschwein_BE <- read_delim("wildschwein_BE_2056.csv", ",")

# convert it to a spatial object
wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), 
                           crs = 2056, 
                           remove = FALSE)
```

# Task 2 - Getting an overview

Calculate the time difference between subsequent rows as described in the demo. You can calculate the time difference using the function difftime() in combination with lead().

- the function difftime() has an option units. Set this to secs to get the time difference in seconds
use as.integer() to turn the output returned by difftime() into an integer.

- store the output in a new column (e.g. timelag)

```{r calculate time difference}
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(timelag = as.integer(difftime(lead(DatetimeUTC),DatetimeUTC),
                           units = "secs"))

```

Now inspect your data in more detail. Try to answer the following questions:

- How many individuals were tracked?

```{r individuals}
wildschwein_BE_count <- wildschwein_BE %>%
  group_by(TierID) %>%
  count()
```

There are three individuals that were tracked. They have the IDs 002A, 0016A, and 018A.

- For how long were the individual tracked? Are there gaps? Were all individuals tracked concurrently or sequentially?

```{r}
ggplot(wildschwein_BE, aes(DatetimeUTC, TierName)) + 
  geom_point()

```

- What is the temporal sampling interval between the locations?

```{r plot histogram}
# timelag/60 = minutes

ggplot(wildschwein_BE, aes(timelag/60)) + 
  geom_histogram(binwidth = 0.5) + 
  lims(x = c(0, 5000/60))+
  scale_y_log10()
```

```{r}
wildschwein_BE %>%
  filter(DatetimeUTC < "2014-08-24") %>%
  ggplot(aes(DatetimeUTC, timelag, color = TierName)) + 
  geom_point() + 
  geom_line()

```

# Task 3 - Deriving movement parameters I: Speed

```{r}
N1 <- 1204752
N2 <- 1204863
E1 <- 2570409
E2 <- 2570402

sqrt((E1-E2)^2 + (N1-N2)^2)

```


```{r}
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(steplength_m = sqrt((E-lead(E))^2 + (N-lead(N))^2))
```

```{r}
wildschwein_BE <- wildschwein_BE %>% 
  mutate(speed_ms = steplength_m/timelag)
```

```{r}

hist(wildschwein_BE$speed_ms, 200)
```

